---
title: 'Statistical Inference on Global Health Data: A comprehensive analysis of the WHO Life Expectancy Dataset'
author: 'Wesley M. Gathua | Reg No: 223347'
date: "2026-06-19"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    extra_dependencies:
    - longtable
    - fvextra
    keep_tex: true
    toc: false
fontsize: 10pt
#bibliography: references.bib
#csl: apa.csl
header-includes:
- \usepackage{microtype}
- \usepackage{fvextra}
- \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
- \sloppy
geometry: left=2cm, right=2cm, top=2cm, bottom=2cm
---

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

required_pkgs <- c("ggplot2", "corrplot", "patchwork", "kableExtra", "broom", "stargazer")

new_pkgs <- required_pkgs[!required_pkgs %in% rownames(installed.packages())]
if (length(new_pkgs) > 0) {
  install.packages(new_pkgs, repos = "[https://cloud.r-project.org](https://cloud.r-project.org)")
}

library(ggplot2)
library(corrplot)
library(patchwork)
library(kableExtra)
library(broom)
library(stargazer)
library(janitor)
library(tidyverse)

# Global ggplot theme
theme_set(theme_minimal(base_size = 12) +
            theme(plot.title = element_text(face = "bold", size = 13),
                  plot.subtitle = element_text(color = "grey40")))
```

```{r loading the data, echo=FALSE}
df <- read.csv("C:/Users/HP/OneDrive - Strathmore University/STRATH UNI/FIRST YEAR/SEMESTER III/UNITS/STATISTICAL INFERENCE IN BIG DATA/STATISTICAL_INFERENCE_(BIG_DATA)/STATISTICAL-INFERENCE-PROJECT/Life-Expectancy-Data-Updated.csv")
```
The dataset was loaded successfully. We proceeded to do some checks onto the dataset.\

```{r data investigation, echo=TRUE}
# Checking for dimensions
cat("Dimensions:", dim(df), "\n\n")

# Checking column names
names(df)

# Checking data types
str(df)

# Checking for null values
colSums(is.na(df))
```
We saw that there 2864 rows and 21 columns in the dataset.\
There were no missing values in all the variables. In order to be sure, we also proceeded to check if there were rows labeled 'null'.

```{r investigating masked missing values, echo = TRUE}
# Checking if there were values masked as 'null', 'NULL' or " "
cat("Searching for text based masked nulls \n")
colSums(df == "null" | df == "NULL" | df == "", na.rm = TRUE)

# Checking for suspicious zeros (0) in the health metrics
cat("\n Investigating null values masquerading as 0s \n")

cat("Zeros in Life Expectancy:", sum(df$Life_expectancy == 0, na.rm = TRUE), "\n" )
cat("Zeros in under 5 deaths:", sum(df$Under_five_deaths == 0, na.rm = TRUE), "\n")
cat("Zeros in under Adult Mortality:", sum(df$Adult_mortality == 0, na.rm = TRUE), "\n")
cat("Zeros in under alcohol consumption:", sum(df$Alcohol_consumption == 0, na.rm = TRUE), "\n")
cat("Zeros in under BMI:", sum(df$BMI == 0, na.rm = TRUE), "\n")
cat("Zeros in under Life Expectancy:", sum(df$Life_expectancy == 0, na.rm = TRUE), "\n")
cat("Zeros in under GDP per capita:", sum(df$GDP_per_capita == 0, na.rm = TRUE), "\n")
cat("Zeros in incidents in HIV:", sum(df$Incidents_HIV == 0, na.rm = TRUE), "\n")
```
It was seen that there were no masked values as "null", 'null' or an empty 
string.\
Upon investigating the zeros, there was no suspicious activity of zeros. The
selected columns for this check were guided by real world scenarios. This is
such as no country could have a GDP of zero or an entire population could have 
a BMI of zero. We did not check for the various diseases because it is possible 
for the various developed countries to have incidences where there are no 
certain diseases because of their well developed health care system which can
also be backed by their robust immunization procedures. With regards with
alcohol where we obtained 38 zeros, we needed to identify if these countries 
came from the Muslim majority countries or countries under an Islamic regime 
of which according to their religious views, consumption of alcohol is
prohibited.\

```{r alcohol investigation, echo = TRUE}
# Checking which countries have an alcohol consumption of zero
unique(df$Country[df$Alcohol_consumption == 0])

```
The Muslim majority hypothesis proved to be correct as the countries with 0
alcohol consumption came from Islamic nations.\
We needed to check if the highly-developed countries indeed had no 
vaccine preventable diseases. We also proceeded to investigate if presence 
of zeros meant that no immunization coverage was done or it meant that there
were no such cases.\

```{r disease investigations, echo = TRUE}
# Checking if it was immunization coverage or case count
cat("Summary of disease columns \n")
summary(df[ , c("Measles", "Polio", "Diphtheria")])
```
An inspection of these diseases proved that there was a strict maximum threshold
of 99. For actual cases, this value would scale to higher figures such as tens
of thousands. This evidently proved that these values were actual percentage
coverage rather than actual case counts.\
We needed to do an investigation of these percentages. This was to check whether
the percentages meant 99\% covered or not covered. This was done by the use of 
a correlation. If it meant covered, there would be an inverse relationship
between the disease and `Infant_deaths` or `Under_five_deaths`.\

```{r vaccination coverage investigation, echo = TRUE}
cat("The correlation between the Measles disease (coverage) and death of
    children under 5 years is", cor(df$Measles, df$Under_five_deaths, use = 'complete.obs'), "\n\n")

cat("The correlation between the Polio disease (coverage) and death of
    infants is", cor(df$Polio, df$Infant_deaths, use = 'complete.obs'), "\n\n")
```
This showed an inverse relationship between the diseases and the child mortality.
This meant that the values exhibited by the diseases meant the extensiveness of
the vaccinations. A 99\% coverage meant that most parts of this particular
country was vaccinated against the specific disease.\

The project's next step was to check if indeed the high immunization coverage
came from the highly developed countries. This would validate our hypothesis of
developed countries having better healthcare. We began by investigating the 
robustness of `Economy_status_Developing` and `Economy_status_Developed`
features to ensure that they were mutually exclusive.\

```{r economy checks, echo = TRUE}
# Investigating that the two economy columns do not add up to 2. Mutual
# exclusivity is needed
economy_check <- df$Economy_status_Developing + df$Economy_status_Developed

# Summary of the calculation
cat("Summary of the economy checks \n")
summary(economy_check)

# Tabular form of these two columns
cat("\n Cross validation table of the two economy features \n")
table(Developed = df$Economy_status_Developed, Developing = df$Economy_status_Developing)
```
We saw that the two columns were indeed mutually exclusive as they both did not
add up to 2. The cross validation table backs this up whereby there are no
individual countries where they are both developed and developing. This is given by the
`0`. Additionally, the table also informs us that there are `2272` records from 
developing countries and there are `592` records from developed countries.\

We now proceeded to see if the high vaccination coverage came from developed or
developing countries.\

```{r vaccination by economy, echo=TRUE}


```

